### 2023-04-26 chapter1/chapter1.tex
Moved citations closer to relevant phrases

Given this importance, methods for inferring orthologous groups of proteins were developed shortly after the first genomes were sequenced in the late 1990s~\cite{Fleischmann1995, Goffeau1996, CESC1998}. One early and influential approach was to cluster triangles of hits resulting from homology searches between pairs of genomes~\cite{Tatusov1997}. Graph-based approaches have remained popular, and in the intervening years many other researchers have refined this method by implementing various pre- and post-processing steps. Despite these improvements, many databases and pipelines use the same triangle clustering algorithm or other methods which require relatively few hits between sequences to infer an orthologous group, \textit{e.g.} connected components or Markov clustering~\cite{Remm2001, Enright2002, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}. However, the scale of biological sequence data has changed dramatically. For example, in the last decade, the number of annotated genomes available from NCBI has increased nearly 20-fold and currently exceeds 900 (Fig.~\ref{sfig:ncbi}). Though this figure is only a rough proxy of the total number of assemblies available, it will likely continue to grow rapidly in the coming years as many large-scale genome assembly efforts such as i5K, the Bird 10,000 Genomes Project, and the Vertebrate Genomes Project have already yielded results~\cite{Thomas2020, Feng2020, Rhie2021}. Thus, the dense taxonomic sampling made possible by these projects poses new challenges and opportunities for the standard methods of orthology inference and alignment, which implicitly assume fewer and more distantly related genomes or fail to fully leverage the redundancy and organization of their annotations.

@@

The identification of sets of densely connected nodes in graphs is known as community detection in network analysis. While many community detection algorithms are available, only some are commonly used in the context of orthology inference. One early method that remains popular is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster~\cite{Tatusov1997, Jensen2007}. Other approaches include the MCL algorithm, which clusters graphs by simulating stochastic flow, and connected components or other single-linkage criteria~\cite{Remm2001, Enright2002, Li2003, Emms2015, Train2017, Cosentino2018}. While these methods are robust when clustering hit graphs derived from smaller or more diverse sets of genomes, they are not suitable for the large number of closely related genomes in this work since they require relatively few edges to define a cluster. For example, the MCL algorithm and connected components method assign a node to a cluster as long as it has a single edge, and triangle clustering only requires two edges to two adjacent nodes.

@@

The most popular aligners for protein sequences (Clustal Omega, MAFFT, MUSCLE, T-Coffee) do not include splice sites in their alignments, which makes them prone to aligning non-homologous exons~\cite{Sievers2017, Katoh2013, Edgar2004, Notredame2000}. Current algorithms can easily be extended to incorporate splice sites by coding them as a new symbol and preventing alignment between splice sites and amino acids, which was recently implemented in the aligner, \textit{Mirage}~\cite{Nord2018}. The biggest challenge in practice, however, is mapping a protein sequence to its genomic sequence to identify splice sites in the protein sequence. Although this information can in principle be derived from the GTF annotation files produced by the NCBI pipeline, annotating the splice junctions in the protein sequences themselves would facilitate splice-sensitive alignment.

@@

Though benchmarks are available for optimizing and comparing methods of orthology inference, the metrics are calculated over a set of reference genomes which are sparsely sampled over a broad taxonomic range, so it is unclear if they are informative for method designed to yield robust inferences when the genomes are highly related~\cite{Nevers2022}. Furthermore, the heterogeneity of genome architectures and annotations may require quality assurance methods tailored to each set of genomes. Thus, there is likely no one-size-fits-all approach to orthology inference, and with many other standalone programs available for more standard use cases (Hieranoid, OMA standalone, OrthoFinder, OrthoInspector, Orthologer), we have chosen not to package the code into an end-to-end pipeline~\cite{Kaduk2017, Altenhoff2019, Emms2019, Linard2014, Zdobnov2020}. Instead, we have devoted considerable attention to organizing and documenting the code to make it accessible to a newcomer and thereby facilitate the adaptation of specific steps to similar projects as needed.

@@

Though databases of orthologous groups such as COGs, Ensembl Compara, EggNOG, OMA, OrthoDB, OrthoInspector, and OrthoMCL will continue to be useful for comparative studies across broad taxonomic ranges, the increasing speed at which high-quality genome assemblies and annotations are produced means no single database can encompass the most complete data~\cite{Galperin2020, Herrero2016, HuertaCepas2018, Altenhoff2020, Zdobnov2020, Nevers2018, Chen2006}. Furthermore, since many early comparative genomics studies spanned diverse branches of the tree of life, future research will likely prioritize taxonomic depth over breadth. Thus, custom sets of orthologous groups will grow more and not less common. Despite the challenges these developments pose, they also present new opportunities to bridge the gap between mutational and macroevolutionary processes.

@@

\subsection{Inference of species trees}
The orthology inference pipeline was first repeated with the outgroup species \textit{Scaptodrosophila lebanonensis}. Orthologous groups with one sequence for each species were aligned, and 100 meta-alignments were constructed by randomly sampling 10,000 columns from these 9,435 alignments. (The alignments were not refined before sampling.) To determine the effect of invariant columns and gaps, two sampling strategies were used where invariant columns were allowed or disallowed and the maximum fraction of gaps was set at 0, 50, and 100\%. Their combination yielded six different sets of meta-alignments. A tree was fit to each meta-alignment with the LG substitution model, four discrete gamma rate categories, and optimized state frequencies using IQ-TREE 1.6.12~\cite{Nguyen2014, Le2008, Yang1994}. If the sampling strategy allowed invariant columns, an invariant rate category was included. The resulting trees from each set were merged into a majority consensus tree (Fig.~\ref{sfig:trees_LG}). All figures are derived from the maximum 50\% gap fraction meta-alignment set unless otherwise noted.

@@

\subsection{Code and data availability}
\begin{sloppypar}
The code used to produce the results and analyses is available at \url{https://github.com/marcsingleton/orthology_inference2023}. HMM algorithms were implemented in the standalone package Homomorph which is available at \url{https://github.com/marcsingleton/homomorph} and on the Python Package Index (PyPI). The following Python libraries were used: matplotlib, NumPy, pandas, and SciPy~\cite{Hunter2007, Harris2020, McKinney2010, Virtanen2020}. Relevant output files are available in the supporting information. There is no primary data associated with this manuscript. All primary data are available from publicly accessible sources described in their corresponding sections.
\end{sloppypar}

>>>>

Given this importance, methods for inferring orthologous groups of proteins were developed shortly after the first genomes were sequenced in the late 1990s~\cite{Fleischmann1995, Goffeau1996, CESC1998}. One early and influential approach was to cluster triangles of hits resulting from homology searches between pairs of genomes~\cite{Tatusov1997}. Graph-based approaches have remained popular, and in the intervening years many other researchers have refined this method by implementing various pre- and post-processing steps. Despite these improvements, many databases and pipelines use the same triangle clustering algorithm~\cite{Jensen2007} or other methods which require relatively few hits between sequences to infer an orthologous group, \textit{e.g.} connected components and other single-linkage criteria~\cite{Remm2001, Train2017, Linard2011, Cosentino2018} or Markov clustering~\cite{Enright2002, Li2003, Emms2015}. However, the scale of biological sequence data has changed dramatically. For example, in the last decade, the number of annotated genomes available from NCBI has increased nearly 20-fold and currently exceeds 900 (Fig.~\ref{sfig:ncbi}). Though this figure is only a rough proxy of the total number of assemblies available, it will likely continue to grow rapidly in the coming years as many large-scale genome assembly efforts such as i5K~\cite{Thomas2020}, the Bird 10,000 Genomes Project~\cite{Feng2020}, and the Vertebrate Genomes Project~\cite{Rhie2021} have already yielded results. Thus, the dense taxonomic sampling made possible by these projects poses new challenges and opportunities for the standard methods of orthology inference and alignment, which implicitly assume fewer and more distantly related genomes or fail to fully leverage the redundancy and organization of their annotations.

@@

The identification of sets of densely connected nodes in graphs is known as community detection in network analysis. While many community detection algorithms are available, only some are commonly used in the context of orthology inference. One early method that remains popular is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster~\cite{Tatusov1997, Jensen2007}. Other approaches include the MCL algorithm, which clusters graphs by simulating stochastic flow~\cite{Enright2002, Li2003, Emms2015}, and connected components or other single-linkage criteria~\cite{Remm2001, Linard2011, Train2017, Cosentino2018}. While these methods are robust when clustering hit graphs derived from smaller or more diverse sets of genomes, they are not suitable for the large number of closely related genomes in this work since they require relatively few edges to define a cluster. For example, the MCL algorithm and connected components method assign a node to a cluster as long as it has a single edge, and triangle clustering only requires two edges to two adjacent nodes.

@@

The most popular aligners for protein sequences (Clustal Omega~\cite{Sievers2017}, MAFFT~\cite{Katoh2013}, MUSCLE~\cite{Edgar2004}, T-Coffee~\cite{Notredame2000}) do not include splice sites in their alignments, which makes them prone to aligning non-homologous exons. Current algorithms can easily be extended to incorporate splice sites by coding them as a new symbol and preventing alignment between splice sites and amino acids, which was recently implemented in the aligner, \textit{Mirage}~\cite{Nord2018}. The biggest challenge in practice, however, is mapping a protein sequence to its genomic sequence to identify splice sites in the protein sequence. Although this information can in principle be derived from the GTF annotation files produced by the NCBI pipeline, annotating the splice junctions in the protein sequences themselves would facilitate splice-sensitive alignment.

@@

Though benchmarks are available for optimizing and comparing methods of orthology inference, the metrics are calculated over a set of reference genomes which are sparsely sampled over a broad taxonomic range, so it is unclear if they are informative for method designed to yield robust inferences when the genomes are highly related~\cite{Nevers2022}. Furthermore, the heterogeneity of genome architectures and annotations may require quality assurance methods tailored to each set of genomes. Thus, there is likely no one-size-fits-all approach to orthology inference, and with many other standalone programs available for more standard use cases (Hieranoid~\cite{Kaduk2017}, OMA standalone~\cite{Altenhoff2019}, OrthoFinder~\cite{Emms2019}, OrthoInspector~\cite{Linard2014}, Orthologer~\cite{Zdobnov2020}), we have chosen not to package the code into an end-to-end pipeline. Instead, we have devoted considerable attention to organizing and documenting the code to make it accessible to a newcomer and thereby facilitate the adaptation of specific steps to similar projects as needed.

@@

Though databases of orthologous groups such as COGs~\cite{Galperin2020}, Ensembl Compara~\cite{Herrero2016}, EggNOG~\cite{HuertaCepas2018}, OMA~\cite{Altenhoff2020}, OrthoDB~\cite{Zdobnov2020}, OrthoInspector~\cite{Nevers2018}, and OrthoMCL~\cite{Chen2006} will continue to be useful for comparative studies across broad taxonomic ranges, the increasing speed at which high-quality genome assemblies and annotations are produced means no single database can encompass the most complete data. Furthermore, since many early comparative genomics studies spanned diverse branches of the tree of life, future research will likely prioritize taxonomic depth over breadth. Thus, custom sets of orthologous groups will grow more and not less common. Despite the challenges these developments pose, they also present new opportunities to bridge the gap between mutational and macroevolutionary processes.

@@

\subsection{Inference of species trees}
The orthology inference pipeline was first repeated with the outgroup species \textit{Scaptodrosophila lebanonensis}. Orthologous groups with one sequence for each species were aligned, and 100 meta-alignments were constructed by randomly sampling 10,000 columns from these 9,435 alignments. (The alignments were not refined before sampling.) To determine the effect of invariant columns and gaps, two sampling strategies were used where invariant columns were allowed or disallowed and the maximum fraction of gaps was set at 0, 50, and 100\%. Their combination yielded six different sets of meta-alignments. A tree was fit to each meta-alignment with the LG substitution model~\cite{Le2008}, four discrete gamma rate categories~\cite{Yang1994}, and optimized state frequencies using IQ-TREE 1.6.12~\cite{Nguyen2014}. If the sampling strategy allowed invariant columns, an invariant rate category was included. The resulting trees from each set were merged into a majority consensus tree (Fig.~\ref{sfig:trees_LG}). All figures are derived from the maximum 50\% gap fraction meta-alignment set unless otherwise noted.

@@

\subsection{Code and data availability}
\begin{sloppypar}
The code used to produce the results and analyses is available at \url{https://github.com/marcsingleton/orthology_inference2023}. HMM algorithms were implemented in the standalone package Homomorph which is available at \url{https://github.com/marcsingleton/homomorph} and on the Python Package Index (PyPI). The following Python libraries were used: matplotlib~\cite{Hunter2007}, NumPy~\cite{Harris2020}, pandas~\cite{McKinney2010}, and SciPy~\cite{Virtanen2020}. Relevant output files are available in the supporting information. There is no primary data associated with this manuscript. All primary data are available from publicly accessible sources described in their corresponding sections.
\end{sloppypar}

### 2023-04-26 chapter2/chapter2.tex
Moved citations closer to relevant phrases

\subsection{Code and data availability}
\begin{sloppypar}
The code used to produce the results and analyses is available at \url{https://github.com/marcsingleton/IDR_evolution2023}. The following Python libraries were used: matplotlib, NumPy, pandas, SciPy, and scikit-learn~\cite{Hunter2007, Harris2020, McKinney2010, Virtanen2020, Pedregosa2011}. Relevant output files are available in the supporting information. There is no primary data associated with this manuscript. All primary data are available from publicly accessible sources described in their corresponding sections.
\end{sloppypar}

>>>>

\subsection{Code and data availability}
\begin{sloppypar}
The code used to produce the results and analyses is available at \url{https://github.com/marcsingleton/IDR_evolution2023}. The following Python libraries were used: matplotlib~\cite{Hunter2007}, NumPy~\cite{Harris2020}, pandas~\cite{McKinney2010}, SciPy~\cite{Virtanen2020}, and scikit-learn~\cite{Pedregosa2011}. Relevant output files are available in the supporting information. There is no primary data associated with this manuscript. All primary data are available from publicly accessible sources described in their corresponding sections.
\end{sloppypar}
